Method to infer relatedness using low coverage ancient DNA

### 2022-02-18 paper_overleaf/main.tex

Even though this may not hold true when we bin sites in large windows, we still find that $Var(X_{x_0}) > Var(X_{x_1}) > Var(X_{x_2})$ (fig. \ref{FigS1:binom}).

@@

To test KIn on real ancient dataset, we applied it to Neandertal specimens from Chagyrskaya cave in Siberia, Russia. This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}. These specimens showed low coverage of nuclear genome, with 8 samples out of 14 showing $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena. All these samples were captured with an array targeting for Neanderthal variation, and we further ascertained this data to variable sites in high coverage Neanderthal genomes: Altai Neanderthal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. This dataset had all the issues that KIn deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_kin}. We found identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relatives (Chagyrskaya07-Chagyrskaya17, Chagyrskaya06-Chagyrskaya14), and a second degree relative (Chagyrskaya01-Chagyrskaya60). These results agree with READ (fig. S\ref{figS2:Chagyrskaya_READ}). Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, and KIn further identifies them as parent-child. Our result matches with the conclusions of the authors based on additional information from mtDNA. Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60). We compared our results to lcMLkin, and show that the software does not perform well with these low coverage samples since it is not possible to obtain reliable genotype likelihoods for many of these samples (fig. S\ref{figS6:Chagyrskaya_ibd}).

@@

We further applied KIn to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019. We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree. Out of 6903 pairs, 346 pairs show different relatedness for KIn and READ (table S2). However, there are only 31 pairs where KIn has $log likelihood >1$. We looked in detail at some of these pairs, and we found that it is possible to deduct the correct relatedness for several of these pairs by plotting the proportion of differences along the genome. In all the pairs that we analyzed in this way, KIn seems to predict the true relatedness. We show few examples of such cases in fig. S\ref{figS8:eg1}. We found 20 pairs in this subset of 346 pairs where lcMLkin has sufficient data. Among these 20 pairs, we found that lcMLkin matches KIn estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIn predicts second degree) and pair AITI72-AITI77A (READ and lcMLKin predict unrelated, while KIn estimates 3rd degree with log likelihood=1.7).  
We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLkin and KIn for these cases (suppl table3). There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIn. Among the 19 overlapping pairs, we found four pairs where KIn and lcMLkin differ, and a visual inspection shows that in these cases, KIn seems to have the correct estimate (fig. S\ref{figS9:eg2}).

@@

Several methods are available to estimate relatedness from modern as well as ancient DNA data. We develop a new method to estimate genetic kinship from low coverage data. We show that it is possible to accurately infer relatedness, along with IBD tracts, from such data in presence of long ROH, contamination, as well as ascertainment bias. Our method utilizes Hidden Markov Models to estimate IBD tracts, conditional on a particular relatedness, and we use log likelihoods of these HMMs to find the best fit. This approach of using Hidden Markov Model makes use of the fact that DNA is inherited in blocks, and hence the IBD state in a given window is dependent on the IBD state in the previous window. We use this approach to identify the following relatives: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals. We show that it is possible to differentiate between avuncular and grand-parent-grandchild relations. We have developed a similar Hidden Markov Model to infer long ROH, and use the posterior probabilities to make better relatedness inference. We further use a simple model to correct the pairwise differences in genomic windows for contamination.

We demonstrate using simulations, that our method performs well at low coverage. We compared KIn's model performance to a widely used software READ to show that our approach works at least as good as READ on ideal data without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives. We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate. Contrary to the intuition, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIn and READ. READ's performance improves because ROH does not cause any change in pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals. More is the relatedness, the more is the decrease in pairwise differences in presence of ROH. And so long ROH causes higher power for READ which otherwise is a conservative method. KIn performs well for the same reasons, but can miss-classify second degree individuals with ROH as siblings, if ROH is not accounted for. Hence, we implement ROH states in the relatedness model to avoid such miss classifications. Contamination, on the other hand can make the performance of READ much worse depending on the amount of contamination, and the number of individuals contaminated. We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls dramatically. Contamination in more individuals will cause a shift in median difference in the population and may cause non contaminated individuals to look more related, thus increasing the false positive rates. On the other hand, increase in amount of contamination in few individuals can make them look more unrelated. KIn, with the implementation of a model to account for contamination performs well in this case. Ascertainment causes a decrease in power for both the methods, since it reduces the amount of data. However, the decrease in power is more for READ compared to KIn.

DNA data from Bronze age, Lech valley has low contamination, and no ROH. We demonstrate that KIn matches the results from READ and lcMLkin for pairs with good amount of overlaps, while adds more information for low coverage samples. As expected from the simulations, KIn is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative. We show that when applied to neandertal specimens from Chagyrskaya cave, READ and KIn agree in relatedness estimates for all pairs where both methods have sufficient data. KIn additionally identifys a pair of first degree relatives as parent-child, which is also confirmed with mtDNA. In addition, KIn identifies a third degree relative. We compare IBD estimates from KIn to lcMLkin and show that KIn's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIn. 
Output of KIn is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio. Due to the tabular output, KIn is easy to automatise and apply on large datasets. 
One limitation with our approach is that this method assumes a single population. In case of population structure, with the subpopulations with very different diversity, KIn may show inaccurate inferences. Our method makes another assumption, that the dataset comprises of at least some unrelated individuals. We estimate $p_0$ as the median of all pairwise differences, and if more than half of the pairs are related to each other, $p_0$ estimate would be biased, and may affect the relatedness inference. In this study, we identify six different cases of relatedness, but in theory our approach can be extended to identify many more such cases, such as double first cousins, using a corresponding IBD state transition matrix.

>>>>

Even though this may not hold true when we bin sites in large windows, we still find that $Var(X_{x_0}) > Var(X_{x_1}) > Var(X_{x_2})$ (fig. \ref{figS1:binom}).

@@

To test KIn on real ancient dataset, we applied it to Neandertal specimens from Chagyrskaya cave in Siberia, Russia. This dataset comprised of 14 specimens that came from neighbouring layers, and hence could possibly belong to contemporary neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{kolobova_archaeological_2020-1}. These specimens showed low coverage of nuclear genome, with 8 samples out of 14 showing $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena. All these samples were captured with an array targeting for Neanderthal variation, and we further ascertained this data to variable sites in high coverage Neanderthal genomes: Altai Neanderthal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. This dataset had all the issues that KIn deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_kin}. We found identical individuals (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), first degree relative (Chagyrskaya07-Chagyrskaya17), and a second degree relative (Chagyrskaya01-Chagyrskaya60). These results agree with READ (fig. S\ref{figS2:Chagyrskaya_READ}). Chagyrskaya06-Chagyrskaya14 were estimated to be First Degree by both the methods, but have been classified as Identical pair in the original study, since Chagyrskaya06 is a tooth sample that fits in mandible which is Chgyrskaya14. We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively). Interestingly, for the pair of Chagyrskaya07-Chagyrskaya17, READ shows that they are first degree relatives, and KIn further identifies them as parent-child. Our result matches with the conclusions of the authors based on additional information from mtDNA. Further, we find a possible third degree relative (Chagyrskaya07-Chagyrskaya60). We compared our IBD estimates to lcMLkin for all pairs for whom READ results (with s.d. >1) and KIn results ($\Lambda$>1) match (fig. S\ref{figS6:Chagyrskaya_ibd}). Chagyrskaya06-Chagyrskaya14 is removed from the analysis. We find that lcMLkin does create four different clusters for different relatedness. However, the clusters are quite deviated from the expected values, which makes it difficult to identify the relatedness. Identical individuals, for example, are expected to be at $k_0 = 0$ and $r = 1$, and Parent-Child at $k_0 = 0$ and $r = 0.5$. These deviations can be due to ascertainment, contamination, ROH as well as low coverage of samples.

@@

We further applied KIn to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019. We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree. Out of 6903 pairs, 346 pairs show different relatedness for KIn and READ (table S2). However, there are only 31 pairs where KIn has $\Lambda >1$. We looked in detail at some of these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome. In all the pairs that we analyzed in this way, KIn seems to predict the true relatedness. We show few examples of such cases in fig. S\ref{figS8:eg1}. We found 20 pairs in this subset of 346 pairs where lcMLkin has sufficient data (Overlapping SNPs $>$ 10000). Among these 20 pairs, we found that lcMLkin matches KIn estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIn predicts second degree) and pair AITI72-AITI77A (READ and lcMLKin predict unrelated, while KIn estimates 3rd degree with log likelihood=1.7).  
We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLkin and KIn for these cases (suppl table3). There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIn. Among the 19 overlapping pairs, we found four pairs where KIn and lcMLkin differ, and a visual inspection shows that in these cases, KIn seems to have the correct estimate (fig. S\ref{figS9:eg2}).

@@

Several methods are available to estimate relatedness from modern as well as ancient DNA data. We developed a new method to estimate genetic kinship from low coverage data. We show that it is possible to accurately infer relatedness, along with IBD tracts, from such data in presence of long ROH, contamination, as well as ascertainment bias. Our method utilizes Hidden Markov Models to estimate IBD tracts, conditional on a particular relatedness, and we use log likelihoods of these HMMs to find the best fit. This approach of using Hidden Markov Model makes use of the fact that DNA is inherited in blocks, and hence the IBD state in a given window is dependent on the IBD state in the previous window. We use this approach to identify the following relatives: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals. We have developed a similar Hidden Markov Model to infer long ROH, and we use the posterior ROH probabilities to make better relatedness inference. We further use a simple model to correct the pairwise differences in genomic windows for contamination.

We demonstrate using simulations, that our method performs well at low coverage. We compared KIn's model performance to a widely used software READ to show that our approach gives similar predictions as READ on ideal data with 4x coverage, and without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives. We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate. Contrary to the intuition, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIn and READ. READ's performance improves because ROH does not cause any change in pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals. More is the relatedness, the more is the decrease in pairwise differences in presence of ROH. And so long ROH causes higher power for READ which otherwise is a conservative method. KIn performs well for the same reasons, but can additionally differentiate between different combinations of IBD and ROH in genomic windows. Contamination, on the other hand can make the performance of READ much worse depending on the amount of contamination, and the number of individuals contaminated. We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls dramatically. Contamination in more individuals will cause a shift in median difference in the population and may cause non contaminated individuals to look more related, thus increasing the false positive rates. On the other hand, increase in amount of contamination in few individuals can make them look more unrelated. KIn, with the implementation of a model to correct for contamination performs well in this case. Ascertainment causes a decrease in power for both the methods, since it reduces the amount of data. However, the decrease in power is more for READ compared to KIn.

DNA data from Bronze age, Lech valley has low contamination, and no ROH. We demonstrate that KIn matches the results from READ and lcMLkin for pairs with good amount of overlaps, while adds more information for low coverage samples. As expected from the simulations, KIn is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative. We show that when applied to Neandertal specimens from Chagyrskaya cave, READ and KIn agree in relatedness estimates for all pairs where both methods have sufficient data. KIn additionally identifys a pair of first degree relatives as parent-child, which is in agreement with mtDNA analysis from the authors. In addition, KIn identifies a third degree relative. We compare IBD estimates from KIn to lcMLkin and show that KIn's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIn. 
Output of KIn is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio. Due to the tabular output, KIn is easy to automatise and apply to large datasets. 
One limitation with our approach is that it assumes a single population. In case of population structure, where the subpopulations have very different diversity, KIn may show inaccurate inferences. Our method makes another assumption, that the dataset comprises of at least some unrelated individuals. We estimate $p_0$ as the median of all pairwise differences, and if more than half of the pairs are related to each other, $p_0$ estimate would be biased, and may affect the relatedness inference. In this study, we identify six different cases of relatedness, but in theory our approach can be extended to identify many more such cases, such as double first cousins, using a corresponding IBD state transition matrix.
